Supporting Information: Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-01-22 paper/current/main.tex
addressing reviewer2 minor comments

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense and continuous vector-space representation, from which we can not only derive a meaningful functional distance between locations but also probe relationships based on culture, language, and even prestige along with the geographic relationship.
We embed trajectories from three massive datasets: U.S. passenger flight itinerary records, Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records.

@@

The axes can be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reason about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Their estimated prestige is speculative, though we find that it significantly correlates with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We also find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations.
Up to a point (around 1,000 researchers), the size of U.S. organization's vectors tends to increase proportionally to the number of researchers (both mobile and non-mobile) with published work;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking.
However, at around 1,000 researchers, the size of the vector \emph{decreases} as the number of researchers increases.
These organizations are primarily research-intensive and prestigious universities with higher rank, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, in which the size of word vectors were found to represent the word's \emph{specificity}, i.e., the word associated with the vector frequently co-appears with particular context words\autocite{schakel2015measuring}.
If the word in question is universal, appearing frequently in many different contexts, it would not have a large norm due to a lack of strong association with a particular context.
Likewise, an organization with a small norm, such as Harvard, appears in many contexts alongside many different organizations in affiliation trajectories---it is well-connected.
The concavity of the curve emerges in part from the relationship between the size of the vector and the expected connectedness of the organization, given its size ($R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector norms because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint.
Some universities, such as the University of Alaska at Fairbanks, have considerably small vectors, which may be a result of their remote locations and unique circumstances.

@@

We define two axes to capture geography and academic prestige, respectively.
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts.
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Ranking or based on the mean normalized research impact sourced from the Leiden Ranking.
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 top-ranked universities are selected and six of them are in the Northeastern U.S., then the bottom twenty will be chosen to also include six from the Northeastern U.S..
From the prestige axis, we derive a ranking of universities that we then compare to other formal university rankings using Spearman rank correlation.

>>>>

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense and continuous vector-space representation, from which we can not only derive a meaningful functional distance between locations but also probe relationships based on culture, language, and even prestige along with the geographic relationship.
We embed trajectories from three massive datasets: U.S. passenger flight itinerary records, Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records (Detailed descriptions are available in the Methods).

@@

The axes can be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reason about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities (Fig.~\ref{fig:semaxis}d), research institutes (Fig.~\ref{fig:semaxis}e), and government organizations (Fig.~\ref{fig:semaxis}f).
Their estimated prestige is speculative, though we find that it significantly correlates with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We also find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations (Fig.~\ref{fig:length}).
Up to a point (around 1,000 researchers), the size of U.S. organization's vectors tends to increase proportionally to the number of researchers (both mobile and non-mobile) with published work;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking.
However, at around 1,000 researchers, the size of the vector \emph{decreases} as the number of researchers increases.
These organizations are primarily research-intensive and prestigious universities with higher rank, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, in which the size of word vectors were found to represent the word's \emph{specificity}, i.e., the word associated with the vector frequently co-appears with particular context words\autocite{schakel2015measuring}.
If the word in question is universal, appearing frequently in many different contexts, it would not have a large norm due to a lack of strong association with a particular context.
Likewise, an organization with a small norm, such as Harvard, appears in many contexts alongside many different organizations in affiliation trajectories---it is well-connected.
The concavity of the curve emerges in part from the relationship between the size of the vector and the expected connectedness of the organization, given its size ($R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector norms because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint.
Some universities, such as the University of Alaska at Fairbanks, have considerably small vectors, which may be a result of their remote locations and unique circumstances.

@@

We define two axes to capture geography and academic prestige, respectively.
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts.
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Ranking or based on the mean normalized research impact sourced from the Leiden Ranking.
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 top-ranked universities are selected and six of them are in the Northeastern U.S., then the bottom twenty will be chosen to also include six from the Northeastern U.S.
From the prestige axis, we derive a ranking of universities that we then compare to other formal university rankings using Spearman rank correlation.

### 2021-01-22 paper/current/supp.tex
fixed swapped links in supp
addressing reviewer2 minor comments

For each mobile researcher who has at least two distinct affiliations, we construct an affiliation trajectory based on the affiliations listed on their published papers indexed in the Web of Science database between 2008 and 2019.
An author is considered mobile if they published with at least two distinct affiliations during the time period of study.
Affiliation names were manually disambiguated, and each was mapped to a unique organization identifier.
An affiliation trajectory for an individual researcher is a sequence of organizations in ascending order of year of publication.
if a researcher published papers with affiliation $A$ in year $t$, $B$ in $t+1$, $C$ in $t+2$ and $A$ again in $t+3$, then the affiliation trajectory is expressed as $(A, B, C, A)$.

@@

%
% Figure - Hyperparameter performance
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Descriptive/hyperparameter_performance.pdf}
	\caption{
		\textbf{Larger dimensions, smaller window size improves embedding performance.}
		The correlation, or amount of flux explained by the embedding distance with varying skip-gram negative sampling hyperparameters.
		Window size refers to $w$, the size of the context window that defines the context in a trajectory.
		Smaller window sizes result in an embedding that explain more flux.
		Embedding dimensions refer to the size of the embedding vector.
		Larger vectors perform better, though little difference between 200 and 300.
		All variants perform better on same-country pairs of organizations than on all organizations.
		All variants perform worse on different-country pairs on organizations.
		Embeddings with larger dimensions outperform mid-size embeddings for the different-country case.
	}
	\label{fig:supp:cosdot}
\end{figure}

%
% Figure - Dot product and cosine similarity
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/d300_ws1_dot_cosine.pdf}
	\caption{
		\textbf{Cosine distance is correlated with dot product similarity.}
		We find a relatively high correlation between the embedding distance---one minus the cosine similarity---and the dot product similarity between organization vectors ($R^2 = 0.73$).
		Color of each hex bin indicates the frequency of organization pairs. 
		Black dots indicate the mean dot product similarity averaged over binned sets at the same embedding distance.
		Red line is line of best fit.
	}
	\label{fig:supp:hyperparams}
\end{figure}

>>>>

For each mobile researcher who has at least two distinct affiliations, we construct an affiliation trajectory based on the affiliations listed on their published papers indexed in the Web of Science database between 2008 and 2019.
An author is considered mobile if they published with at least two distinct affiliations during the time period of study.
Affiliation names were manually disambiguated, and each was mapped to a unique organization identifier.
An affiliation trajectory for an individual researcher is a sequence of organizations in ascending order of year of publication.
If a researcher published papers with affiliation $A$ in year $t$, $B$ in $t+1$, $C$ in $t+2$ and $A$ again in $t+3$, then the affiliation trajectory is expressed as $(A, B, C, A)$.

@@

%
% Figure - Hyperparameter performance
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Descriptive/hyperparameter_performance.pdf}
	\caption{
		\textbf{Larger dimensions, smaller window size improves embedding performance.}
		The correlation, or amount of flux explained by the embedding distance with varying skip-gram negative sampling hyperparameters.
		Window size refers to $w$, the size of the context window that defines the context in a trajectory.
		Smaller window sizes result in an embedding that explain more flux.
		Embedding dimensions refer to the size of the embedding vector.
		Larger vectors perform better, though little difference between 200 and 300.
		All variants perform better on same-country pairs of organizations than on all organizations.
		All variants perform worse on different-country pairs on organizations.
		Embeddings with larger dimensions outperform mid-size embeddings for the different-country case.
	}
	\label{fig:supp:hyperparams}
\end{figure}

%
% Figure - Dot product and cosine similarity
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/d300_ws1_dot_cosine.pdf}
	\caption{
		\textbf{Cosine distance is correlated with dot product similarity.}
		We find a relatively high correlation between the embedding distance---one minus the cosine similarity---and the dot product similarity between organization vectors ($R^2 = 0.73$).
		Color of each hex bin indicates the frequency of organization pairs. 
		Black dots indicate the mean dot product similarity averaged over binned sets at the same embedding distance.
		Red line is line of best fit.
	}
	\label{fig:supp:cosdot}
\end{figure}
